---
title: "BCG case"
bibliography: "references.bib"
output:
  html_document:
    code_folding: show
    code_download: true
---
This pipeline is intended for use with three levels of proteomics data: total proteome, phosphoproteome, and histone proteome. See @Schaefer2024 for more detail on the total and histone proteome; phosphoproteome data is from @Choudhary2020. Total and phosphoproteome data are designed to be input as an Excel sheet from Proteome Discoverer, and the histone data are the aggregated ratios from EpiProfile (@Yuan2018).

The full R Markdown source document can be downloaded from this page by clicking the Source link above. Additionally, we have made available the files generated from our data at the project repository on Github. Other files can be downloaded from their respective sources.

```{r knit, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
if (!requireNamespace('BiocManager', quietly = TRUE)) {
    install.packages('BiocManager')
}
library('BiocManager')

# Install from CRAN if necessary
CRAN_pkgs <- c("stringr", "tibble", "ggvenn", "devtools",
               "reticulate", "igraph", "tidyr", "BiocManager",
               "ggrepel", "scales", "htmltools", "dplyr", "plotthis")
package.check <- sapply(
  CRAN_pkgs,
  FUN = function(x) {
    if (!requireNamespace(x, quietly = TRUE)) {
      install.packages(x, dependencies = TRUE)
      }
    library(x, character.only = TRUE)
  }
)

# Install from Bioconductor if necessary
Bioc_pkgs <- c("clusterProfiler", "enrichplot", "qvalue",
               "OrganismDbi", "Homo.sapiens", "msigdbr")
package.check <- sapply(
  Bioc_pkgs,
  FUN = function(x) {
    if (!requireNamespace(x, quietly = TRUE)) {
      BiocManager::install(x, dependencies = TRUE)
      }
    library(x, character.only = TRUE)
  }
)

library(MPOInt)
remove(Bioc_pkgs, CRAN_pkgs, package.check)
```

Most variables that need to be adjusted for individual datasets should be set here.

```{r set-vars}
total_p_threshold <- 0.05
total_FC_threshold <- 0.5
phospho_p_threshold <- 0.05
phospho_FC_threshold <- 0.5
histone_p_threshold <- 0.05
histone_FC_threshold <- 0.5
```

If you have not previously set up a Python environment from RStudio, run this chunk.

```{r python-1st-setup}
virtualenv_create("r-reticulate")
use_virtualenv("r-reticulate")
virtualenv_install("r-reticulate", c("pandas", "networkx", "pyvis"), action = "add")
```

# 1. Proteome data processing
This section prepares proteome data. For data other than the example, read in the data in place of the `data()` call.

```{r total-processing}
data(THP1_total)
# Renaming to reader-friendly names
THP1_total <- dplyr::rename(THP1_total, all_of(c(
  "log2(FC)" = "Abundance Ratio (log2): (JMI) / (Control)",
  "P value" = "Abundance Ratio Adj. P-Value: (JMI) / (Control)"
)))
# Mapping ENTREZ IDs and a gene symbol column to UniProt IDs for later analysis
THP1_total <- dplyr::left_join(THP1_total, OrganismDbi::select(
  Homo.sapiens::Homo.sapiens,
  keys = OrganismDbi::keys(Homo.sapiens::Homo.sapiens, keytype = "UNIPROT"),
  columns = c("ENTREZID", "SYMBOL"), keytype = "UNIPROT"
), by = dplyr::join_by(Accession == UNIPROT), keep = FALSE)
# Dropping anything with an undefined or duplicate symbol
THP1_total <- THP1_total %>%
  tidyr::drop_na(SYMBOL) %>%
  dplyr::distinct(SYMBOL, .keep_all = TRUE)
# Assigning row names
THP1_total <- THP1_total %>%
  dplyr::mutate("SYMBOL2" = `SYMBOL`) %>%
  tibble::column_to_rownames(var = "SYMBOL") %>% 
  dplyr::rename("Symbol" = "SYMBOL2")
# Calculating Storey's q-value for FDR estimation
THP1_total$`Q value` <- qvalue(THP1_total$`P value`)$qvalues

# Creating a separate set of significant points for later analysis
THP1_total_sig <- dplyr::filter(THP1_total, `Q value` < total_p_threshold)
THP1_total_sig_unadj <- dplyr::filter(THP1_total, `P value` < total_p_threshold)
```

## 1.1. Volcano plots
Here we create several volcano plots to show the data.

```{r total-opts}
# Setting this value - geom_text_repel will only show so many labels at once
# for legibility reasons. Adjust this threshold to show more/fewer points.
options("ggrepel.max.overlaps" = 20)

volcanoplot(df = THP1_total, fc_col = "log2(FC)",
            p_col = "Q value", label_col = "Symbol") +
  ggtitle("Volcano plot") + xlab("log2(FC)") + ylab("-log10(Q-value)")
```

If you have known proteins of interest, use `volcano_groups()` to highlight them on a graph.

```{r total-groups}
# For known groups of interest
volcano_groups(df = THP1_total, fc_col = "log2(FC)", p_col = "Q value", label_col = "Symbol",
  group_list = list("Group 1: Epigenome" = c("MORF4L1", "SAMSN1", "DTX3L", "HSF1",
  "EYA3", "NCAPD2", "TAF9"), "Group 2: Infection" = c("NFKB1", "BST2", "CTSH", "TLR2", "ITGAM", "MMP9", "MX1", "MX2", "C1QBP", "HLA-DRA"))) +
  scale_color_manual(values = c("Group 1: Epigenome" = "red", "Group 2: Infection" = "blue", "None" = "gray")) +
  ggtitle("Grouped volcano plot") + xlab("log2FC") + ylab("-log10(Q-value)")
```

## 1.2. GSEA
This section prepares and performs a GSEA analysis on the total proteome data using the clusterProfiler package.

```{r total-GSEA-list}
# Need a named list in decreasing order
total_GSEA_list <- THP1_total$`log2(FC)`
names(total_GSEA_list) <- THP1_total$ENTREZID
total_GSEA_list <- na.omit(total_GSEA_list)
total_GSEA_list <- sort(total_GSEA_list, decreasing = TRUE)
```

```{r total-enrich}
total_GSEA_GO <- gseGO(geneList = total_GSEA_list, OrgDb = "org.Hs.eg.db",
                    verbose = FALSE, ont = "ALL")
total_GSEA_KEGG <- gseKEGG(geneList = total_GSEA_list, verbose = FALSE)
```

Plotting GSEA dot plots can be difficult due to the volume of information included. Changing the value of `label_format` adjusts the label wrapping, which can help crowded figures.

```{r total-enrich-plots, fig.dpi = 300}
if (nrow(total_GSEA_GO@result) > 0) {
  dotplot(total_GSEA_GO, showCategory = 20, font.size = rel(1), label_format = 40, title = "Total GO enrichment",
          split = "ONTOLOGY") + facet_wrap(ONTOLOGY~.sign)
} else {
  print("Error: no terms enriched under P-value cutoff (total_GSEA_GO)")
}

if (nrow(total_GSEA_KEGG@result) > 0) {
  dotplot(total_GSEA_KEGG, showCategory = 20, font.size = rel(1), label_format = 40, title = "Total KEGG enrichment",
          split = ".sign") + facet_wrap(.~.sign)
} else {
  print("Error: no terms enriched under P-value cutoff (total_GSEA_KEGG)")
}
```
```{r clear-1}
# Clearing large variables
rm(THP1_total, total_GSEA_GO, total_GSEA_KEGG, total_GSEA_list)
```

# 2. Phosphoproteome data processing
Similar to the previous section, this prepares phosphoproteome data.

```{r phosphoproteome-processing}
# Renaming to reader-friendly names not necessary here
data(THP1_phospho)
# Mapping ENTREZ IDs and a gene symbol column to UniProt IDs for later analysis
THP1_phospho <- dplyr::left_join(THP1_phospho, OrganismDbi::select(
  Homo.sapiens::Homo.sapiens,
  keys = OrganismDbi::keys(Homo.sapiens::Homo.sapiens, keytype = "ENTREZID"),
  columns = c("UNIPROT", "SYMBOL"), keytype = "ENTREZID"
), by = dplyr::join_by(SYMBOL2 == SYMBOL, ENTREZID), keep = FALSE, relationship = "many-to-many")
# Dropping anything with an undefined or duplicate symbol
THP1_phospho <- THP1_phospho %>%
  tidyr::drop_na(SYMBOL2) %>%
  dplyr::distinct(SYMBOL2, .keep_all = TRUE)
# Assigning row names
THP1_phospho <- THP1_phospho %>%
  dplyr::mutate("SYMBOL" = `SYMBOL2`) %>%
  tibble::column_to_rownames(var = "SYMBOL") %>% 
  dplyr::rename("Symbol" = "SYMBOL2")
THP1_phospho$`Q value` <- qvalue(THP1_phospho$`P value`)$qvalues

# Creating a separate set of significant points for later analysis
THP1_phospho_sig_unadj <- dplyr::filter(THP1_phospho, `P value` < phospho_p_threshold)
THP1_phospho_sig <- dplyr::filter(THP1_phospho, `Q value` < phospho_p_threshold)
```

## 2.1. Volcano plots
```{r phospho-plot}
# Setting this value - geom_text_repel will only show so many labels at once
# for legibility reasons. Adjust this threshold to show more/fewer points.
options("ggrepel.max.overlaps" = 20)

volcanoplot(df = THP1_phospho, fc_col = "log2(FC)",
            p_col = "Q value", label_col = "Symbol") +
  ggtitle("Volcano plot") + xlab("log2(FC)") + ylab("-log10(Q-value)")
```

```{r phospho-groups}
# For known groups of interest
volcano_groups(df = THP1_phospho, fc_col = "log2(FC)", p_col = "Q value", label_col = "Symbol",
  group_list = list("Group 1: Chromatin" = c("ASH2L", "ASXL1", "ASXL2", "BAP1", "BMI1", "CHD4", "EZH2", "DMAP1",
  "EP400", "KANSL1", "KANSL3", "KMT2D", "MBD3",
  "MEN1", "MGA", "MRGBP", "MTA1", "NCOA6", "REST",
  "RING1", "RUVBL1", "SAP130", "SIRT3", "SIRT6", "EP300"), "Group 2: NMIBC" = c("FANCI", "CDK14", "ICAM1", "CDK1", "NOTCH3", "CASP8",
  "BST2", "CTSL", "MX2", "CASP1", "MX1"))) +
  scale_color_manual(values = c("Group 1: Chromatin" = "red", "Group 2: NMIBC" = "blue", "None" = "gray")) +
  ggtitle("Grouped volcano plot") + xlab("log2FC") + ylab("-log10(Q-value)")
```

## 2.2. GSEA
```{r phospho-GSEA-list}
# Need a named list in decreasing order
phospho_GSEA_list <- THP1_phospho$`log2(FC)`
names(phospho_GSEA_list) <- THP1_phospho$ENTREZID
phospho_GSEA_list <- na.omit(phospho_GSEA_list)
phospho_GSEA_list <- sort(phospho_GSEA_list, decreasing = TRUE)
```

Here we use the `simplify` function from clusterProfiler to combine redundant GSEA terms by similarity. This allows more unique pathways to be displayed.

```{r phospho-enrich}
phospho_GSEA_GO <- gseGO(geneList = phospho_GSEA_list, OrgDb = "org.Hs.eg.db",
                    verbose = FALSE, ont = "ALL")
# By visual inspection of graphs, this contained redundant terms
simple_phospho_GSEA_GO <- clusterProfiler::simplify(phospho_GSEA_GO)
phospho_GSEA_KEGG <- gseKEGG(geneList = phospho_GSEA_list, verbose = FALSE)
```

```{r phospho-enrich-plots}
if (nrow(simple_phospho_GSEA_GO@result) > 0) {
  dotplot(simple_phospho_GSEA_GO, showCategory = 20, font.size = rel(1), label_format = 50, title = "Phospho GO enrichment",
          split = "ONTOLOGY") + facet_wrap(ONTOLOGY~.sign, scales = "free_y", ncol = 2)
} else {
  print("Error: no terms enriched under P-value cutoff (simple_phospho_GSEA_GO)")
}

if (nrow(phospho_GSEA_KEGG@result) > 0) {
  dotplot(phospho_GSEA_KEGG, showCategory = 20, font.size = rel(1), label_format = 40, title = "Phospho KEGG enrichment",
          split = ".sign") + facet_wrap(.~.sign, scales = "free", ncol = 2)
} else {
  print("Error: no terms enriched under P-value cutoff (phospho_GSEA_KEGG)")
}
```

```{r clear-2}
# Clearing large variables
rm(phospho_GSEA_GO, phospho_GSEA_KEGG, simple_phospho_GSEA_GO, THP1_phospho,
   phospho_GSEA_list)
```

# 3. Histone processing
This section processes EpiProfile output into a more readable, user-friendly format. We also perform basic statistical analysis and log transformations on the data.

```{r fragment-histone-processing-names}
data(THP1_histones)
histone_single_raw <- clean_names(THP1_histones, "PTM")[[1]]
histone_frag_raw <- clean_names(THP1_histones, "PTM")[[2]]
```

We separate out the histone fragments containing co-occurring PTMs from the individual PTMs and present both individually. This allows for discovery with complexes that require two or more targets, but also provides a clear picture of overall PTM changes. Here, we group our two conditions together, marked by columns that start with "I" for infected cells or "C" for control.

## 3.1. Fragments
```{r fragment-histone-processing}
# Get the log2-transformed fold change and BH-adjusted P values
histone_frag <- histone_epi(df = histone_frag_raw, label_col = "PTM",
                           base_cols = c("C1", "C2", "C3", "C4", "C5"),
                           target_cols = c("I1", "I2", "I3", "I4", "I5"))
```

The `volcano_epi()` function plots the epiproteome data, assuming that all values have already been transformed and statistical analysis has already been performed. To address infinite fold change values that appear when a protein is not detected in one condition, it also sets them to have a unique shape in the final plot and assigns a fold change value equal to 2.5 times the maximum valid value.

```{r fragment-plot}
# Setting this value - geom_text_repel will only show so many labels at once
# for legibility reasons. Adjust this threshold to show more/fewer points.
options("ggrepel.max.overlaps" = 20)

volcano_epi(df = histone_frag, fc_col = "log2(FC)",
            p_col = "P value", label_col = "PTM", p_thr = 0.1) +
  labs(title = "Volcano plot", x = "log2(FC)", y = "-log10(P-value)")
```

## 3.2. Single
This section parses the fragments into single PTMs. Each occurrence is counted individually - for example, H4 4-17 K5acK12acK16ac would be counted as H4K5ac, H4K8ac, and H4K16ac.

```{r single-processing}
histone_single <- histone_epi(df = histone_single_raw, label_col = "PTM",
                           base_cols = c("C1", "C2", "C3", "C4", "C5"),
                           target_cols = c("I1", "I2", "I3", "I4", "I5"))
```

```{r histone-single-opts}
# Setting this value - geom_text_repel will only show so many labels at once
# for legibility reasons. Adjust this threshold to show more/fewer points.
options("ggrepel.max.overlaps" = 20)

volcano_epi(df = histone_single, fc_col = "log2(FC)",
            p_col = "P_unadj", label_col = "PTM", p_thr = 0.05) +
  labs(title = "Volcano plot", x = "log2(FC)", y = "-log10(P-value)") 
```

```{r clear-3}
# Clearing large variables
rm(THP1_histones, histone_single_raw, histone_frag_raw)
```

# 4. Disease enrichment
Here we perform disease enrichment using the [MSigDB database](https://www.gsea-msigdb.org/gsea/msigdb/human/genesets.jsp) and the MSigDBR package. First, identify a dataset of interest in the database. The next chunk shows databases sourced through MSigDB but originating from WikiPathways, Human Phenotype Ontology, and Reactome. The `collection` parameter identifies the MSigDB collection to search, followed by `gs_exact_source` that specifies the selected dataset. This can also be replaced with `gs_id` if you are using the "Systematic name" value on the website.

We enrich our primary disease of interest (bladder cancer), but also include enrichment for melanoma and RSV, two potential areas of BCG application.

```{r GDA-setup}
GDA_BlCa <- msigdbr(collection = "C2") %>% filter(gs_exact_source == "WP2828")
GDA_melanoma <- msigdbr(collection = "C5") %>% filter(gs_exact_source == "HP:0002861")
GDA_RSV <- msigdbr(collection = "C2") %>% filter(gs_exact_source == "R-HSA-9833110")
```

This chunk creates Venn diagrams of the overlap between proteins that were significant and above the fold change threshold in the total proteome and phosphoproteome, as well as the genes of interest that were just retrieved.

```{r GDA, fig.asp = 1}
# Get lists of proteins that were significant AND above the FC threshold
total_FC <- filter(THP1_total_sig, abs(`log2(FC)`) > total_FC_threshold)
phospho_FC <- filter(THP1_phospho_sig, abs(`log2(FC)`) > phospho_FC_threshold)

# Generate plot
ggvenn(list(
  "NMIBC" = GDA_BlCa$gene_symbol,
  "Total proteome" = total_FC$Symbol,
  "Phosphoproteome" = phospho_FC$Symbol),
  fill_color = c("#440154", "#21908C", "#FDE725"),
  show_percentage = FALSE, text_size = 5) + ggtitle("BCG enrichment") + theme(
  plot.title.position = "plot",
  plot.title = element_text(hjust = 0.5))

ggvenn(list(
  "Melanoma" = GDA_melanoma$gene_symbol,
  "Total proteome" = total_FC$Symbol,
  "Phosphoproteome" = phospho_FC$Symbol),
  fill_color = c("#440154", "#21908C", "#FDE725"),
  show_percentage = FALSE, text_size = 5) + ggtitle("BCG enrichment") + theme(
  plot.title.position = "plot",
  plot.title = element_text(hjust = 0.5))

ggvenn(list(
  "RSV" = GDA_RSV$gene_symbol,
  "Total proteome" = total_FC$Symbol,
  "Phosphoproteome" = phospho_FC$Symbol),
  fill_color = c("#440154", "#21908C", "#FDE725"),
  show_percentage = FALSE, text_size = 10) + ggtitle("BCG enrichment") + theme(
  plot.title.position = "plot",
  plot.title = element_text(hjust = 0.5))
```

We next print a list of proteins in the four intersecting areas for further investigation in our disease of interest.

```{r GDA-list, results = "asis"}
# Print the list of proteins in each intersection
BlCa_intersections <- data.frame("Intersection" = character(),
                                 "Hits" = character(),
                                 stringsAsFactors = FALSE)
BlCa_intersections <- BlCa_intersections %>%
  add_row("Intersection" = "Bladder cancer/Phospho",
          "Hits" = paste(sort(intersect(GDA_BlCa$gene_symbol, phospho_FC$Symbol)),
                         collapse = ", ")) %>% 
  add_row("Intersection" = "Bladder cancer/Total",
          "Hits" = paste(sort(intersect(GDA_BlCa$gene_symbol, total_FC$Symbol)),
                         collapse = ", ")) %>% 
  add_row("Intersection" = "Total/Phospho",
          "Hits" = paste(sort(intersect(total_FC$Symbol, phospho_FC$Symbol)),
                         collapse = ", ")) %>% 
  add_row("Intersection" = "Bladder cancer/Phospho/Total",
          "Hits" = paste(sort(intersect(intersect(
            GDA_BlCa$gene_symbol, phospho_FC$Symbol),
            total_FC$Symbol)), collapse = ", "))

knitr::kable(BlCa_intersections, caption = "Bladder cancer intersections")
```

```{r GDA-list-melanoma, results = "asis", collapse = TRUE}
melanoma_intersections <- data.frame("Intersection" = character(),
                                 "Hits" = character(),
                                 stringsAsFactors = FALSE)
melanoma_intersections <- melanoma_intersections %>%
  add_row("Intersection" = "Melanoma/Phospho",
          "Hits" = paste(sort(intersect(GDA_melanoma$gene_symbol,
                                        phospho_FC$Symbol)),
                         collapse = ", ")) %>% 
  add_row("Intersection" = "Melanoma/Total",
          "Hits" = paste(sort(intersect(GDA_melanoma$gene_symbol,
                                        total_FC$Symbol)),
                         collapse = ", ")) %>% 
  add_row("Intersection" = "Total/Phospho",
          "Hits" = paste(sort(intersect(total_FC$Symbol, phospho_FC$Symbol)),
                         collapse = ", ")) %>% 
  add_row("Intersection" = "Melanoma/Phospho/Total",
          "Hits" = paste(sort(intersect(intersect(
            GDA_melanoma$gene_symbol, phospho_FC$Symbol),
            total_FC$Symbol)), collapse = ", "))

knitr::kable(melanoma_intersections, caption = "Melanoma intersections")
```

```{r GDA-list-RSV, results = "asis", collapse = TRUE}
RSV_intersections <- data.frame("Intersection" = character(),
                                 "Hits" = character(),
                                 stringsAsFactors = FALSE)
RSV_intersections <- RSV_intersections %>%
  add_row("Intersection" = "RSV/Phospho",
          "Hits" = paste(sort(intersect(GDA_RSV$gene_symbol, phospho_FC$Symbol)),
                         collapse = ", ")) %>% 
  add_row("Intersection" = "RSV/Total",
          "Hits" = paste(sort(intersect(GDA_RSV$gene_symbol, total_FC$Symbol)),
                         collapse = ", ")) %>% 
  add_row("Intersection" = "Total/Phospho",
          "Hits" = paste(sort(intersect(total_FC$Symbol, phospho_FC$Symbol)),
                         collapse = ", ")) %>% 
  add_row("Intersection" = "RSV/Phospho/Total",
          "Hits" = paste(sort(intersect(intersect(
            GDA_RSV$gene_symbol, phospho_FC$Symbol),
            total_FC$Symbol)), collapse = ", "))

knitr::kable(RSV_intersections, caption = "RSV intersections")
```



```{r clear-4}
# Clearing large variables
rm(GDA_BlCa, GDA_melanoma, GDA_RSV, BlCa_intersections, melanoma_intersections, RSV_intersections)
```


# 5. Epigenetic network
This section develops a network visualization of all epigenetic interactions as defined by EpiFactors (https://epifactors.autosome.org/). Increased dataset size may make networks difficult to visualize, so an interactive network is drawn in Python and rendered as an embedded HTML file.

The `data(genes)` call imports a .csv file included with the MPOInt package, but can be replaced to update or change the database by downloading and importing the "Table of all proteins" file on the EpiFactors website.

It is important to note that here we choose to use all histones identified and limit the proteome/phosphoproteome by unadjusted P value rather than the FDR-corrected Q value. Given that this tool is intended for discovery and requires downstream validation of its findings, as well as the lack of in-depth resources detailing epigenetic interactions, we chose to include as many potential hits as possible, then evaluate the findings based on the context of the overall developed network in addition to the epiproteome fragment analysis. Below, we also generate a smaller network abiding by more rigorous statistical standards as a demonstration.

## 5.1 Complex and member identification
For future research use, we generate multiple tables containing the identified complexes, members, common targets, and identified PTMs.

```{r network-setup}
# Adding framework for connections
data(genes)
# Creating the network object with the significant peptide subsets
net_created <- 
  multi_net(genes_df = genes, total_df = THP1_total_sig_unadj,
            phospho_df = THP1_phospho_sig_unadj,
            # Point to the gene symbols and transformed FC values
            symbol_col = "Symbol", fc_col = "log2(FC)",
            # The network will not assemble with the fragment data
            epi_single_df = histone_single,
            epi_symbol_col = "PTM", epi_fc_col = "log2(FC)")

# Accessing the plots and lists created as part of the object
net_created$grid
net_created$complex_list
knitr::kable(net_created$member_list)
knitr::kable(net_created$gene_list)
```

```{r tables}
# Getting a dataframe with the significant network members
sn_df <- net_created$ts

# Creating two tables (centered on protein, then centered on histone PTMs) with
# more descriptive information
sn_collapsed_gene <- sn_df %>%
  group_by(Symbol, `log2(FC)_gene`, `P value`) %>%
  summarize(Source = paste(unique(Source), collapse = ", "), PTMs = paste(unique(PTM), collapse = ", "), Complexes = paste(unique(Complex), collapse = ", ")) %>% 
  rename("log2(FC)_gene" = "log2(FC) (gene)")
knitr::kable(sn_collapsed_gene)


sn_collapsed_PTM <- sn_df %>%
  group_by(PTM, `log2(FC)_hist`) %>%
  summarize(Genes = paste(unique(Symbol), collapse = ", "), Complexes = paste(unique(Complex), collapse = ", ")) %>% 
  rename("log2(FC)_hist" = "log2(FC) (PTM)")
knitr::kable(sn_collapsed_PTM)
```

```{r sankey}
# Creating a Sankey flow diagram for an alternate representation
# of how complexes, PTMs, and members are connected
sankey_sn <- sn_df
sankey_sn$Complex <- str_wrap(sankey_sn$Complex, 10)
node_names <- unique(c(sankey_sn$PTM, unique(sankey_sn$Complex)))

v_pal_res <- (viridis_pal(begin = 0.1, option = "D")(length(node_names)))
names(v_pal_res) <- node_names

link_pal <- rep("gray60", times = length(unique(sankey_sn$Symbol)))
names(link_pal) <- unique(sankey_sn$Symbol)

SankeyPlot(arrange(sankey_sn, PTM), x = c("PTM", "Complex"), in_form = "wide",
           links_fill_by = "Symbol", nodes_label = TRUE, xlab = "", ylab = "",
           expand = c(0.1, 0.6), facet_by = "Symbol", facet_scales = "free_y",
           aspect.ratio = 1, facet_ncol = 4, theme_args = list(strip.background = ggplot2::element_rect(fill="grey80")), nodes_legend = "separate",
           links_palcolor = link_pal, links_color = ".fill",
           nodes_palcolor = v_pal_res) +
  theme(legend.position = "none", axis.text.y = element_blank(), axis.ticks.y = element_blank())
```

## 5.2 Python visualization
Finally, we create an interactive network using Python via the `reticulate` package. This allows a dynamic visualization that can be refined and filtered to generate different views.

```{r py-setup}
net_edges_py <- r_to_py(net_created$net_edges)
net_vertices_py <- r_to_py(net_created$net_vertices)
```

```{python py-network-graph}
import pandas as pd, networkx as nx
from pyvis.network import Network
# Create NetworkX graph object
G = nx.Graph()

# Generate a nested dictionary/tuple with the node of interest ("Symbol") and 
# the relevant attributes (shape and complex)
node_list = []
for i, row in r.net_vertices_py.iterrows():
  node_list.append((row["Symbol"], {"color": {"background": row.Colors, "border": row.Border}, "Complex": row.Complex, "shape": row.Shapes}))
G.add_nodes_from(node_list)

# Generate a tuple for edges connecting proteins with PTMs
edge_tup = []
for i in range(len(r.net_edges_py["Symbol"])):
  edge_tup.append((r.net_edges_py["Symbol"][i], r.net_edges_py["PTM"][i]))
G.add_edges_from(edge_tup)

# Remove self-loops
G.remove_edges_from(nx.selfloop_edges(G))

# Now remove nodes without edges
lone_nodes = [node for node, degree in G.degree() if degree == 0]
G.remove_nodes_from(lone_nodes)
# To change the parameters of what's being graphed, adjust the initial call to Network()
net = Network(height="900px", notebook = True, select_menu = True, cdn_resources = "in_line")
net.from_nx(G)
net.save_graph("networkx-pyvis-BCG.html")
```

```{r print-HTML}
htmltools::renderDocument(htmltools::htmlTemplate("networkx-pyvis-BCG.html"))
```

## 5.3 Minimal network

```{r network-setup-2}
# This value (and selected metric) can be adjusted to fit the data
histone_single_p <- filter(histone_single, `P_unadj` < 0.2)

# Now making a network with a subset of histones, as well as
# proteome/phosphoproteome peptides that meet the FDR-adjusted threshold
net_created_hp <- multi_net(genes_df = genes, total_df = THP1_total_sig,
          phospho_df = THP1_phospho_sig, symbol_col = "Symbol", fc_col = "log2(FC)",
          epi_single_df = histone_single_p,
          epi_symbol_col = "PTM", epi_fc_col = "log2(FC)")
net_created_hp$grid
net_created_hp$complex_list
net_created_hp$member_list
net_created_hp$gene_list
net_created_hp$PTM_list
```

```{r tables-2}
# Getting a dataframe with the network members
sn_df_hp <- net_created_hp$ts

# Creating two tables (centered on protein, then centered on histone PTMs) with
# more descriptive information
sn_collapsed_gene_hp <- sn_df_hp %>%
  group_by(Symbol, `log2(FC)_gene`, `P value`) %>%
  summarize(Source = paste(unique(Source), collapse = ", "), PTMs = paste(unique(PTM), collapse = ", "), Complexes = paste(unique(Complex), collapse = ", ")) %>% 
  rename("log2(FC)_gene" = "log2(FC) (gene)")
knitr::kable(sn_collapsed_gene_hp)


sn_collapsed_PTM_hp <- sn_df_hp %>%
  group_by(PTM, `log2(FC)_hist`) %>%
  summarize(Genes = paste(unique(Symbol), collapse = ", "), Complexes = paste(unique(Complex), collapse = ", ")) %>% 
  rename("log2(FC)_hist" = "log2(FC) (PTM)")
knitr::kable(sn_collapsed_PTM_hp)
```



```{r sankey-2}
# Creating a Sankey flow diagram for an alternate representation
# of how complexes, PTMs, and members are connected
sankey_sn_hp <- sn_df_hp
sankey_sn_hp$Complex <- str_wrap(sankey_sn_hp$Complex, 10)
node_names_hp <- unique(c(sankey_sn_hp$PTM, unique(sankey_sn_hp$Complex)))

v_pal_res_hp <- (viridis_pal(begin = 0.1, option = "D")(length(node_names_hp)))
names(v_pal_res_hp) <- node_names_hp

link_pal_hp <- rep("gray60", times = length(unique(sankey_sn_hp$Symbol)))
names(link_pal_hp) <- unique(sankey_sn_hp$Symbol)

SankeyPlot(arrange(sankey_sn_hp, PTM), x = c("PTM", "Complex"), in_form = "wide",
           links_fill_by = "Symbol", nodes_label = TRUE, xlab = "", ylab = "",
           expand = c(0.1, 0.6), facet_by = "Symbol", facet_scales = "free_y",
           aspect.ratio = 1, facet_ncol = 4, theme_args = list(strip.background = ggplot2::element_rect(fill="grey80")), nodes_legend = "separate",
           links_palcolor = link_pal_hp, links_color = ".fill",
           nodes_palcolor = v_pal_res_hp) +
  theme(legend.position = "none", axis.text.y = element_blank(), axis.ticks.y = element_blank())
```

```{r py-setup-2}
net_edges_py_hp <- r_to_py(net_created_hp$net_edges)
net_vertices_py_hp <- r_to_py(net_created_hp$net_vertices)
```

```{python py-network-graph-2}
import pandas as pd, networkx as nx
from pyvis.network import Network
# Create NetworkX graph object
G2 = nx.Graph()

# Generate a nested dictionary/tuple with the node of interest ("Symbol") and 
# the relevant attributes (shape and complex)
node_list_hp = []
for i, row in r.net_vertices_py_hp.iterrows():
  node_list_hp.append((row["Symbol"], {"color": {"background": row.Colors, "border": row.Border}, "Complex": row.Complex, "shape": row.Shapes}))
G2.add_nodes_from(node_list_hp)

# Generate a tuple for edges connecting proteins with PTMs
edge_tup_hp = []
for i in range(len(r.net_edges_py_hp["Symbol"])):
  edge_tup_hp.append((r.net_edges_py_hp["Symbol"][i], r.net_edges_py_hp["PTM"][i]))
G2.add_edges_from(edge_tup_hp)

# Remove self-loops
G2.remove_edges_from(nx.selfloop_edges(G2))

# Now remove nodes without edges
lone_nodes_hp = [node for node, degree in G2.degree() if degree == 0]
G2.remove_nodes_from(lone_nodes_hp)
# To change the parameters of what's being graphed, adjust the initial call to Network()
net2 = Network(height="900px", notebook = True, select_menu = True, cdn_resources = "in_line")
net2.from_nx(G2)
net2.save_graph("networkx-pyvis-BCG2.html")
```

```{r print-HTML-2}
htmltools::renderDocument(htmltools::htmlTemplate("networkx-pyvis-BCG2.html"))
```
