---
title: "MCF7 case"
bibliography: "references.bib"
output:
  html_document:
    code_folding: show
    code_download: true
---
This pipeline is intended for use with three levels of proteomics data: total proteome, phosphoproteome, and histone proteome. Total and phosphoproteome data are sourced from @Han2024. Total and phosphoproteome data are designed to be input as an Excel sheet from MaxQuant, and the histone data are derived from @LeRoy2013.

The full R Markdown source document can be downloaded from this page by clicking the Source link above. Additionally, we have made available the files generated from our data at the project repository on Github. Other files can be downloaded from their respective sources.

```{r knit, include = FALSE}
library(knitr)
library(rmarkdown)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
if (!requireNamespace('BiocManager', quietly = TRUE)) {
    install.packages('BiocManager')
}
library('BiocManager')

# Install from CRAN if necessary
CRAN_pkgs <- c("stringr", "tibble", "ggvenn", "devtools",
               "reticulate", "igraph", "tidyr", "BiocManager",
               "ggrepel", "scales", "htmltools", "dplyr", "plotthis")
package.check <- sapply(
  CRAN_pkgs,
  FUN = function(x) {
    if (!requireNamespace(x, quietly = TRUE)) {
      install.packages(x, dependencies = TRUE)
      }
    library(x, character.only = TRUE)
  }
)

# Install from Bioconductor if necessary
Bioc_pkgs <- c("clusterProfiler", "enrichplot", "qvalue",
               "OrganismDbi", "Homo.sapiens", "msigdbr")
package.check <- sapply(
  Bioc_pkgs,
  FUN = function(x) {
    if (!requireNamespace(x, quietly = TRUE)) {
      BiocManager::install(x, dependencies = TRUE)
      }
    library(x, character.only = TRUE)
  }
)

library(MPOInt)
remove(Bioc_pkgs, CRAN_pkgs, package.check)
```

Most variables that need to be adjusted for individual datasets should be set here.

```{r set-vars}
total_p_threshold <- 0.05
total_FC_threshold <- 0.5
phospho_p_threshold <- 0.05
phospho_FC_threshold <- 0.5
histone_p_threshold <- 0.05
histone_FC_threshold <- 0.5
```

If you have not previously set up a Python environment from RStudio, run this chunk.

```{r python-1st-setup}
virtualenv_create("r-reticulate")
use_virtualenv("r-reticulate")
virtualenv_install("r-reticulate", c("pandas", "networkx", "pyvis"), action = "add")
```

# 1. Proteome data processing
This section prepares proteome data. For data other than the example, read in the data in place of the `data()` call.

```{r total-processing}
data(MCF7_total)
# Renaming to reader-friendly names not necessary here
# Mapping ENTREZ and UNIPROT IDs already done
# Dropping anything with an undefined or duplicate symbol
MCF7_total <- MCF7_total %>%
  tidyr::drop_na(SYMBOL2) %>%
  dplyr::distinct(SYMBOL2, .keep_all = TRUE) %>% 
  dplyr::rename("Symbol" = "SYMBOL2")
# Assigning row names already done
# Calculating Storey's q-value for FDR estimation
MCF7_total$`Q value` <- qvalue(MCF7_total$`P value`)$qvalues

# Creating a separate set of significant points for later analysis
MCF7_total_sig <- dplyr::filter(MCF7_total, `Q value` < total_p_threshold)
MCF7_total_sig_unadj <- dplyr::filter(MCF7_total, `P value` < total_p_threshold)
```

## 1.1. Volcano plots
Here we create several volcano plots to show the data.

```{r total-opts}
# Setting this value - geom_text_repel will only show so many labels at once
# for legibility reasons. Adjust this threshold to show more/fewer points.
options("ggrepel.max.overlaps" = 20)

volcanoplot(df = MCF7_total, fc_col = "log2(FC)",
            p_col = "Q value", label_col = "Symbol") +
  ggtitle("Volcano plot") + xlab("log2(FC)") + ylab("-log10(Q-value)")
```

If you have known proteins of interest, use `volcano_groups()` to highlight them on a graph.

```{r total-groups}
# For known groups of interest
# Setting this value - geom_text_repel will only show so many labels at once
# for legibility reasons. Adjust this threshold to show more/fewer points.
options("ggrepel.max.overlaps" = 20)
volcano_groups(df = MCF7_total, fc_col = "log2(FC)", p_col = "Q value", label_col = "Symbol",
  group_list = list("Group 1: Chromatin" = c("KMT2A", "H1-5", "H2BC21", "EP400", "RBBP6", "KDM4B", "CHD6"), "Group 2: ERBC" = c("RUNX1", "PTEN", "MAP3K1", "CCND1", "MYC", "AHNAK2", "IL20RA", "CCNB1", "EIF4EBP2", "PTGFRN", "PRKDC", "TBL1XR1", "CDKN1B", "SF3B5"))) +
  scale_color_manual(values = c("Group 1: Chromatin" = "red", "Group 2: ERBC" = "blue", "None" = "gray")) +
  ggtitle("Grouped volcano plot") + xlab("log2FC") + ylab("-log10(Q-value)")
```

## 1.2. GSEA
This section prepares and performs a GSEA analysis on the total proteome data.

```{r total-GSEA-list}
# Need a named list in decreasing order
total_GSEA_list <- MCF7_total$`log2(FC)`
names(total_GSEA_list) <- MCF7_total$ENTREZID
total_GSEA_list <- na.omit(total_GSEA_list)
total_GSEA_list <- sort(total_GSEA_list, decreasing = TRUE)
```

```{r total-enrich}
total_GSEA_GO <- gseGO(geneList = total_GSEA_list, OrgDb = "org.Hs.eg.db",
                    verbose = FALSE, ont = "ALL")
# By visual inspection, this contains overlapping terms
simple_total_GSEA_GO <- clusterProfiler::simplify(total_GSEA_GO)
total_GSEA_KEGG <- gseKEGG(geneList = total_GSEA_list, verbose = FALSE)
```

Plotting GSEA dot plots can be difficult due to the volume of information included. Changing the value of `label_format` adjusts the label wrapping, which can help crowded figures.

```{r total-enrich-plots, fig.dpi = 300}
if (nrow(total_GSEA_GO@result) > 0) {
  dotplot(total_GSEA_GO, showCategory = 20, font.size = rel(0.8), label_format = 40, title = "GO enrichment",
          split = "ONTOLOGY") + facet_wrap(ONTOLOGY~.sign, scales = "free_y", ncol = 2)
} else {
  print("Error: no terms enriched under P-value cutoff (total_GSEA_GO)")
}

if (nrow(simple_total_GSEA_GO@result) > 0) {
  dotplot(simple_total_GSEA_GO, showCategory = 20, font.size = rel(1), label_format = 40, title = "Simplified GO enrichment",
          split = "ONTOLOGY") + facet_wrap(ONTOLOGY~.sign, scales = "free_y", ncol = 2)
} else {
  print("Error: no terms enriched under P-value cutoff (simple_total_GSEA_GO)")
}

if (nrow(total_GSEA_KEGG@result) > 0) {
  dotplot(total_GSEA_KEGG, showCategory = 20, font.size = rel(1), label_format = 40, title = "KEGG enrichment",
          split = ".sign") + facet_wrap(.~.sign)
} else {
  print("Error: no terms enriched under P-value cutoff (total_GSEA_KEGG)")
}
```
```{r clear-1}
# Clearing large variables
rm(total_GSEA_GO, simple_total_GSEA_GO, total_GSEA_KEGG, total_GSEA_list, MCF7_total)
```



# 2. Phosphoproteome data processing
Similar to the previous section, this prepares phosphoproteome data.

```{r phosphoproteome-processing}
# Renaming to reader-friendly names not necessary here
data(MCF7_phospho)
# Mapping ENTREZ IDs and a gene symbol column to UniProt IDs for later analysis
MCF7_phospho <- dplyr::left_join(MCF7_phospho, OrganismDbi::select(
  Homo.sapiens::Homo.sapiens,
  keys = OrganismDbi::keys(Homo.sapiens::Homo.sapiens, keytype = "ENTREZID"),
  columns = c("UNIPROT", "SYMBOL"), keytype = "ENTREZID"
), by = dplyr::join_by(SYMBOL2 == SYMBOL, ENTREZID), keep = FALSE, relationship = "many-to-many")
# Dropping anything with an undefined or duplicate symbol
MCF7_phospho <- MCF7_phospho %>%
  tidyr::drop_na(SYMBOL2) %>%
  dplyr::distinct(SYMBOL2, .keep_all = TRUE)
# Assigning row names
MCF7_phospho <- MCF7_phospho %>%
  dplyr::mutate("SYMBOL" = `SYMBOL2`) %>%
  tibble::column_to_rownames(var = "SYMBOL") %>% 
  dplyr::rename("Symbol" = "SYMBOL2")
# Calculating Storey's q-value for FDR estimation
MCF7_phospho$`Q value` <- qvalue(MCF7_phospho$`P value`)$qvalues

# Creating a separate set of significant points for later analysis
MCF7_phospho_sig <- dplyr::filter(MCF7_phospho, `Q value` < phospho_p_threshold)
MCF7_phospho_sig_unadj <- dplyr::filter(MCF7_phospho, `P value` < phospho_p_threshold)
```

## 2.1. Volcano plots
```{r phospho-plot}
# Setting this value - geom_text_repel will only show so many labels at once
# for legibility reasons. Adjust this threshold to show more/fewer points.
options("ggrepel.max.overlaps" = 20)

volcanoplot(df = MCF7_phospho, fc_col = "log2(FC)",
            p_col = "Q value", label_col = "Symbol") +
  ggtitle("Volcano plot") + xlab("log2(FC)") + ylab("-log10(Q-value)")
```

```{r phospho-groups}
# For known groups of interest
volcano_groups(df = MCF7_phospho, fc_col = "log2(FC)", p_col = "Q value", label_col = "Symbol",
  group_list = list("Group 1: Chromatin" = c("ASH2L", "SAP130", "REST", "CHD4", "EED", "KDM4A", "KDM1A", "RBBP4", "HIST1H4A", "KAT7", "SMARCAD1"), "Group 2: ERBC" = c("CDK1", "GATAD2B", "AKT2", "PARP1", "XRCC6", "AHNAK", "MAPK1"))) +
  scale_color_manual(values = c("Group 1: Chromatin" = "red", "Group 2: ERBC" = "blue", "None" = "gray")) +
  ggtitle("Grouped volcano plot") + xlab("log2FC") + ylab("-log10(Q-value)")
```

## 2.2. GSEA
```{r phospho-GSEA-list}
# Need a named list in decreasing order
MCF7_phospho <- drop_na(MCF7_phospho, ENTREZID) %>% distinct(ENTREZID, .keep_all = TRUE)
phospho_GSEA_list <- MCF7_phospho$`log2(FC)`
names(phospho_GSEA_list) <- MCF7_phospho$ENTREZID
phospho_GSEA_list <- sort(phospho_GSEA_list, decreasing = TRUE)
```

```{r phospho-enrich}
phospho_GSEA_GO <- gseGO(geneList = phospho_GSEA_list, OrgDb = "org.Hs.eg.db",
                    verbose = FALSE, ont = "ALL")
phospho_GSEA_KEGG <- gseKEGG(geneList = phospho_GSEA_list, verbose = FALSE)
```

```{r phospho-enrich-plots}
if (nrow(phospho_GSEA_GO@result) > 0) {
  dotplot(phospho_GSEA_GO, showCategory = 20, font.size = rel(1), label_format = 50, title = "GO enrichment",
          split = "ONTOLOGY") + facet_wrap(ONTOLOGY~.sign, scales = "free_y", ncol = 2)
} else {
  print("Error: no terms enriched under P-value cutoff (simple_phospho_GSEA_GO)")
}

if (nrow(phospho_GSEA_KEGG@result) > 0) {
  dotplot(phospho_GSEA_KEGG, showCategory = 20, font.size = rel(1), label_format = 40, title = "KEGG enrichment",
          split = ".sign") + facet_wrap(.~.sign, scales = "free", ncol = 2)
} else {
  print("Error: no terms enriched under P-value cutoff (phospho_GSEA_KEGG)")
}
```

```{r clear-2}
# Clearing large variables
rm(phospho_GSEA_GO, phospho_GSEA_KEGG,
   phospho_GSEA_list, MCF7_phospho)
```

# 3. Histone processing
This section processes EpiProfile output into a more readable, user-friendly format. In this case, the epiproteome data is in a summary statistics format, where each condition is only presented as a mean value with standard deviation. We create a standard baseline and then generate a P value from a T score.

```{r histone-import-mean}
data(MCF7_histones)
MCF7_histones <- dplyr::select(MCF7_histones, c("PTM", "MCF7 ave (5 reps)", "293 ave (5 reps)", "HaCAT ave (5 reps)", "hESC ave (6 reps)", "HFF ave (7 reps)", "Mdm13 ave (5 reps)", "MCF7 std"))
histone_single_raw <- clean_names(MCF7_histones, "PTM")[[1]]
histone_frag_raw <- clean_names(MCF7_histones, "PTM")[[2]]
```

## 3.1. Fragments
The data available here was already in a legible format, so we do not have to process it extensively.

```{r histone-frag-process}
histone_frag <- histone_summary(df = histone_frag_raw, label_col = "PTM", base_cols = c("293 ave (5 reps)", "HaCAT ave (5 reps)", "hESC ave (6 reps)", "HFF ave (7 reps)", "Mdm13 ave (5 reps)"), base_std = NA, target_col = "MCF7 ave (5 reps)", target_std = "MCF7 std")
```

```{r fragment-plot}
# Setting this value - geom_text_repel will only show so many labels at once
# for legibility reasons. Adjust this threshold to show more/fewer points.
options("ggrepel.max.overlaps" = 7)

volcano_epi(df = histone_frag, fc_col = "log2(FC)",
            p_col = "P_adj", label_col = "PTM") +
  labs(title = "Volcano plot", x = "log2(FC)", y = "-log10(P-value)")
```

## 3.2. Single
This section parses the fragments into single PTMs. Each occurrence is counted individually - for example, H4 4-17 K5acK12acK16ac would be counted as H4K5ac, H4K8ac, and H4K16ac.

```{r histone-single-process}
histone_single <- histone_summary(df = histone_single_raw, label_col = "PTM", base_cols = c("293 ave (5 reps)", "HaCAT ave (5 reps)", "hESC ave (6 reps)", "HFF ave (7 reps)", "Mdm13 ave (5 reps)"), base_std = NA, target_col = "MCF7 ave (5 reps)", target_std = "MCF7 std")
```


```{r fragment-plot-2}
# Setting this value - geom_text_repel will only show so many labels at once
# for legibility reasons. Adjust this threshold to show more/fewer points.
options("ggrepel.max.overlaps" = 20)

volcano_epi(df = histone_single, fc_col = "log2(FC)",
            p_col = "P_adj", label_col = "PTM") +
  labs(title = "Volcano plot", x = "log2(FC)", y = "-log10(P-value)")
```

```{r clear-3}
# Clearing large variables
rm(MCF7_histones, histone_frag_raw, histone_single_raw)
```

# 4. Disease enrichment
Here we perform disease enrichment using the [MSigDB database](https://www.gsea-msigdb.org/gsea/msigdb/human/genesets.jsp) and the MSigDBR package. First, identify a dataset of interest in the database. The next chunk shows databases sourced through MSigDB but originating from PubMed and Human Phenotype Ontology. The `collection` parameter identifies the MSigDB collection to search, followed by `gs_exact_source` that specifies the selected dataset. This can also be replaced with `gs_id` if you are using the "Systematic name" value on the website.

We enrich our primary disease of interest (breast cancer), but also include enrichment for hypertension and ovarian cancer.

```{r GDA-setup}
GDA_ERBC <- msigdbr(collection = "C2") %>% filter(gs_id == "M18299")
GDA_HT <- msigdbr(collection = "C5") %>% filter(gs_exact_source == "HP:0000822")
GDA_OvCa <- msigdbr(collection = "C5") %>% filter(gs_exact_source == "HP:0100615")
```

This chunk creates Venn diagrams of the overlap between proteins that were significant and above the fold change threshold in the total and phosphoproteome, as well as the genes of interest that were just retrieved.

```{r GDA, fig.asp = 1}
# Get lists of proteins that were significant AND above the FC threshold
total_FC <- filter(MCF7_total_sig, abs(`log2(FC)`) > total_FC_threshold)
phospho_FC <- filter(MCF7_phospho_sig, abs(`log2(FC)`) > phospho_FC_threshold)

# Generate plot
ggvenn(list(
  "ER breast cancer" = GDA_ERBC$gene_symbol,
  "Total proteome" = total_FC$Symbol,
  "Phosphoproteome" = phospho_FC$Symbol),
  fill_color = c("#440154", "#21908C", "#FDE725"),
  show_percentage = FALSE, text_size = 5) + ggtitle("MCF-7 enrichment") + theme(
  plot.title.position = "plot",
  plot.title = element_text(hjust = 0.5))

ggvenn(list(
  "Hypertension" = GDA_HT$gene_symbol,
  "Total proteome" = total_FC$Symbol,
  "Phosphoproteome" = phospho_FC$Symbol),
  fill_color = c("#440154", "#21908C", "#FDE725"),
  show_percentage = FALSE, text_size = 5) + ggtitle("MCF-7 enrichment") + theme(
  plot.title.position = "plot",
  plot.title = element_text(hjust = 0.5))

ggvenn(list(
  "Ovarian cancer" = GDA_OvCa$gene_symbol,
  "Total proteome" = total_FC$Symbol,
  "Phosphoproteome" = phospho_FC$Symbol),
  fill_color = c("#440154", "#21908C", "#FDE725"),
  show_percentage = FALSE, text_size = 5) + ggtitle("MCF-7 enrichment") + theme(
  plot.title.position = "plot",
  plot.title = element_text(hjust = 0.5))
```

We next print a list of proteins in the four intersecting areas for further investigation in our disease of interest.

```{r GDA-list, results = "asis"}
# Print the list of proteins in each intersection
ERBC_intersections <- data.frame("Intersection" = character(),
                                 "Hits" = character(),
                                 stringsAsFactors = FALSE)
ERBC_intersections <- ERBC_intersections %>%
  add_row("Intersection" = "ER breast cancer/Phospho",
          "Hits" = paste(sort(intersect(GDA_ERBC$gene_symbol, phospho_FC$Symbol)),
                         collapse = ", ")) %>% 
  add_row("Intersection" = "ER breast cancer/Total",
          "Hits" = paste(sort(intersect(GDA_ERBC$gene_symbol, total_FC$Symbol)),
                         collapse = ", ")) %>% 
  add_row("Intersection" = "Total/Phospho",
          "Hits" = paste(sort(intersect(total_FC$Symbol, phospho_FC$Symbol)),
                         collapse = ", ")) %>% 
  add_row("Intersection" = "ER breast cancer/Phospho/Total",
          "Hits" = paste(sort(intersect(intersect(
            GDA_ERBC$gene_symbol, phospho_FC$Symbol),
            total_FC$Symbol)), collapse = ", "))

knitr::kable(ERBC_intersections, caption = "ERBC intersections")
```

```{r GDA-list-HT, results = "asis", collapse = TRUE}
HT_intersections <- data.frame("Intersection" = character(),
                                 "Hits" = character(),
                                 stringsAsFactors = FALSE)
HT_intersections <- HT_intersections %>%
  add_row("Intersection" = "Hypertension/Phospho",
          "Hits" = paste(sort(intersect(GDA_HT$gene_symbol, phospho_FC$Symbol)),
                         collapse = ", ")) %>% 
  add_row("Intersection" = "Hypertension/Total",
          "Hits" = paste(sort(intersect(GDA_HT$gene_symbol, total_FC$Symbol)),
                         collapse = ", ")) %>% 
  add_row("Intersection" = "Total/Phospho",
          "Hits" = paste(sort(intersect(total_FC$Symbol, phospho_FC$Symbol)),
                         collapse = ", ")) %>% 
  add_row("Intersection" = "Hypertension/Phospho/Total",
          "Hits" = paste(sort(intersect(intersect(
            GDA_HT$gene_symbol, phospho_FC$Symbol),
            total_FC$Symbol)), collapse = ", "))

knitr::kable(HT_intersections, caption = "Hypertension intersections")
```

```{r GDA-list-OvCa, results = "asis", collapse = TRUE}
OvCa_intersections <- data.frame("Intersection" = character(),
                                 "Hits" = character(),
                                 stringsAsFactors = FALSE)
OvCa_intersections <- OvCa_intersections %>%
  add_row("Intersection" = "Ovarian cancer/Phospho",
          "Hits" = paste(sort(intersect(GDA_OvCa$gene_symbol, phospho_FC$Symbol)),
                         collapse = ", ")) %>% 
  add_row("Intersection" = "Ovarian cancer/Total",
          "Hits" = paste(sort(intersect(GDA_OvCa$gene_symbol, total_FC$Symbol)),
                         collapse = ", ")) %>% 
  add_row("Intersection" = "Total/Phospho",
          "Hits" = paste(sort(intersect(total_FC$Symbol, phospho_FC$Symbol)),
                         collapse = ", ")) %>% 
  add_row("Intersection" = "Ovarian cancer/Phospho/Total",
          "Hits" = paste(sort(intersect(intersect(
            GDA_OvCa$gene_symbol, phospho_FC$Symbol),
            total_FC$Symbol)), collapse = ", "))

knitr::kable(OvCa_intersections, caption = "Ovarian cancer intersections")
```

```{r clear-4}
# Clearing large variables
rm(GDA_ERBC, GDA_HT, GDA_OvCa, ERBC_intersections, HT_intersections, OvCa_intersections)
```


# 5. Epigenetic network
This section develops a network visualization of all epigenetic interactions as defined by EpiFactors (https://epifactors.autosome.org/). Increased dataset size may make networks difficult to visualize, so an interactive network is drawn in Python and rendered as an embedded HTML file.

The `data(genes)` call imports a .csv file included with the MPOInt package, but can be replaced to update or change the database by downloading and importing the "Table of all proteins" file on the EpiFactors website.

It is important to note that here we choose to use all histones identified and limit the proteome/phosphoproteome by unadjusted P value rather than the FDR-corrected Q value. Given that this tool is intended for discovery and requires downstream validation of its findings, as well as the lack of in-depth resources detailing epigenetic interactions, we chose to include as many potential hits as possible, then evaluate the findings based on the context of the overall developed network in addition to the epiproteome fragment analysis. Below, we also generate a smaller network abiding by more rigorous statistical standards as a demonstration.

## 5.1 Complex and member identification
For future research use, we generate multiple tables containing the identified complexes, members, common targets, and identified PTMs.

```{r network-setup}
# Adding framework for connections
data(genes)
# Creating the network object with the significant peptide subsets
net_created <- 
  multi_net(genes_df = genes, total_df = MCF7_total_sig_unadj,
            phospho_df = MCF7_phospho_sig_unadj,
            # Point to the gene symbols and transformed FC values
            symbol_col = "Symbol", fc_col = "log2(FC)",
            # The network will not assemble with the fragment data
            epi_single_df = histone_single,
            epi_symbol_col = "PTM", epi_fc_col = "log2(FC)")

# Accessing the plots and lists created as part of the object
net_created$grid
net_created$complex_list
knitr::kable(net_created$member_list)
knitr::kable(net_created$gene_list)
```

```{r tables}
# Getting a dataframe with the significant network members
sn_df <- net_created$ts

# Creating two tables (centered on protein, then centered on histone PTMs) with
# more descriptive information
sn_collapsed_gene <- sn_df %>%
  group_by(Symbol, `log2(FC)_gene`, `P value`) %>%
  summarize(Source = paste(unique(Source), collapse = ", "), PTMs = paste(unique(PTM), collapse = ", "), Complexes = paste(unique(Complex), collapse = ", ")) %>% 
  rename("log2(FC)_gene" = "log2(FC) (gene)")
knitr::kable(sn_collapsed_gene)


sn_collapsed_PTM <- sn_df %>%
  group_by(PTM, `log2(FC)_hist`) %>%
  summarize(Genes = paste(unique(Symbol), collapse = ", "), Complexes = paste(unique(Complex), collapse = ", ")) %>% 
  rename("log2(FC)_hist" = "log2(FC) (PTM)")
knitr::kable(sn_collapsed_PTM)
```

```{r sankey}
# Creating a Sankey flow diagram for an alternate representation
# of how complexes, PTMs, and members are connected
sankey_sn <- sn_df
sankey_sn$Complex <- str_wrap(sankey_sn$Complex, 10)
node_names <- unique(c(sankey_sn$PTM, unique(sankey_sn$Complex)))

v_pal_res <- (viridis_pal(begin = 0.1, option = "D")(length(node_names)))
names(v_pal_res) <- node_names

link_pal <- rep("gray60", times = length(unique(sankey_sn$Symbol)))
names(link_pal) <- unique(sankey_sn$Symbol)

SankeyPlot(arrange(sankey_sn, PTM), x = c("PTM", "Complex"), in_form = "wide",
           links_fill_by = "Symbol", nodes_label = TRUE, xlab = "", ylab = "",
           expand = c(0.1, 0.6), facet_by = "Symbol", facet_scales = "free_y",
           aspect.ratio = 1, facet_ncol = 4, theme_args = list(strip.background = ggplot2::element_rect(fill="grey80")), nodes_legend = "separate",
           links_palcolor = link_pal, links_color = ".fill",
           nodes_palcolor = v_pal_res) +
  theme(legend.position = "none", axis.text.y = element_blank(), axis.ticks.y = element_blank())
```

## 5.2 Python visualization
Finally, we create an interactive network using Python via the `reticulate` package. This allows a dynamic visualization that can be refined and filtered to generate different views.

```{r py-setup}
net_edges_py <- r_to_py(net_created$net_edges)
net_vertices_py <- r_to_py(net_created$net_vertices)
```

```{python py-network-graph}
import pandas as pd, networkx as nx
from pyvis.network import Network
# Create NetworkX graph object
G = nx.Graph()

# Generate a nested dictionary/tuple with the node of interest ("Symbol") and 
# the relevant attributes (shape and complex)
node_list = []
for i, row in r.net_vertices_py.iterrows():
  node_list.append((row["Symbol"], {"color": {"background": row.Colors, "border": row.Border}, "Complex": row.Complex, "shape": row.Shapes}))
G.add_nodes_from(node_list)

# Generate a tuple for edges connecting proteins with PTMs
edge_tup = []
for i in range(len(r.net_edges_py["Symbol"])):
  edge_tup.append((r.net_edges_py["Symbol"][i], r.net_edges_py["PTM"][i]))
G.add_edges_from(edge_tup)

# Remove self-loops
G.remove_edges_from(nx.selfloop_edges(G))

# Now remove nodes without edges
lone_nodes = [node for node, degree in G.degree() if degree == 0]
G.remove_nodes_from(lone_nodes)
# To change the parameters of what's being graphed, adjust the initial call to Network()
net = Network(height="900px", notebook = True, select_menu = True, cdn_resources = "in_line")
net.from_nx(G)
net.save_graph("networkx-pyvis-MCF.html")
```

```{r print-HTML}
htmltools::renderDocument(htmltools::htmlTemplate("networkx-pyvis-MCF.html"))
```

## 5.3 Minimal network

```{r network-setup-2}
# This value (and selected metric) can be adjusted to fit the data
histone_single_p <- filter(histone_single, `P_unadj` < 0.2)

# Now making a network with a subset of histones, as well as
# proteome/phosphoproteome peptides that meet the FDR-adjusted threshold
net_created_hp <- multi_net(genes_df = genes, total_df = MCF7_total_sig,
          phospho_df = MCF7_phospho_sig, symbol_col = "Symbol", fc_col = "log2(FC)",
          epi_single_df = histone_single_p,
          epi_symbol_col = "PTM", epi_fc_col = "log2(FC)")
net_created_hp$grid
net_created_hp$complex_list
net_created_hp$member_list
net_created_hp$gene_list
net_created_hp$PTM_list
```

```{r tables-2}
# Getting a dataframe with the network members
sn_df_hp <- net_created_hp$ts

# Creating two tables (centered on protein, then centered on histone PTMs) with
# more descriptive information
sn_collapsed_gene_hp <- sn_df_hp %>%
  group_by(Symbol, `log2(FC)_gene`, `P value`) %>%
  summarize(Source = paste(unique(Source), collapse = ", "), PTMs = paste(unique(PTM), collapse = ", "), Complexes = paste(unique(Complex), collapse = ", ")) %>% 
  rename("log2(FC)_gene" = "log2(FC) (gene)")
knitr::kable(sn_collapsed_gene_hp)


sn_collapsed_PTM_hp <- sn_df_hp %>%
  group_by(PTM, `log2(FC)_hist`) %>%
  summarize(Genes = paste(unique(Symbol), collapse = ", "), Complexes = paste(unique(Complex), collapse = ", ")) %>% 
  rename("log2(FC)_hist" = "log2(FC) (PTM)")
knitr::kable(sn_collapsed_PTM_hp)
```



```{r sankey-2}
# Creating a Sankey flow diagram for an alternate representation
# of how complexes, PTMs, and members are connected
sankey_sn_hp <- sn_df_hp
sankey_sn_hp$Complex <- str_wrap(sankey_sn_hp$Complex, 10)
node_names_hp <- unique(c(sankey_sn_hp$PTM, unique(sankey_sn_hp$Complex)))

v_pal_res_hp <- (viridis_pal(begin = 0.1, option = "D")(length(node_names_hp)))
names(v_pal_res_hp) <- node_names_hp

link_pal_hp <- rep("gray60", times = length(unique(sankey_sn_hp$Symbol)))
names(link_pal_hp) <- unique(sankey_sn_hp$Symbol)

SankeyPlot(arrange(sankey_sn_hp, PTM), x = c("PTM", "Complex"), in_form = "wide",
           links_fill_by = "Symbol", nodes_label = TRUE, xlab = "", ylab = "",
           expand = c(0.1, 0.6), facet_by = "Symbol", facet_scales = "free_y",
           aspect.ratio = 1, facet_ncol = 4, theme_args = list(strip.background = ggplot2::element_rect(fill="grey80")), nodes_legend = "separate",
           links_palcolor = link_pal_hp, links_color = ".fill",
           nodes_palcolor = v_pal_res_hp) +
  theme(legend.position = "none", axis.text.y = element_blank(), axis.ticks.y = element_blank())
```

```{r py-setup-2}
net_edges_py_hp <- r_to_py(net_created_hp$net_edges)
net_vertices_py_hp <- r_to_py(net_created_hp$net_vertices)
```

```{python py-network-graph-2}
import pandas as pd, networkx as nx
from pyvis.network import Network
# Create NetworkX graph object
G2 = nx.Graph()

# Generate a nested dictionary/tuple with the node of interest ("Symbol") and 
# the relevant attributes (shape and complex)
node_list_hp = []
for i, row in r.net_vertices_py_hp.iterrows():
  node_list_hp.append((row["Symbol"], {"color": {"background": row.Colors, "border": row.Border}, "Complex": row.Complex, "shape": row.Shapes}))
G2.add_nodes_from(node_list_hp)

# Generate a tuple for edges connecting proteins with PTMs
edge_tup_hp = []
for i in range(len(r.net_edges_py_hp["Symbol"])):
  edge_tup_hp.append((r.net_edges_py_hp["Symbol"][i], r.net_edges_py_hp["PTM"][i]))
G2.add_edges_from(edge_tup_hp)

# Remove self-loops
G2.remove_edges_from(nx.selfloop_edges(G2))

# Now remove nodes without edges
lone_nodes_hp = [node for node, degree in G2.degree() if degree == 0]
G2.remove_nodes_from(lone_nodes_hp)
# To change the parameters of what's being graphed, adjust the initial call to Network()
net2 = Network(height="900px", notebook = True, select_menu = True, cdn_resources = "in_line")
net2.from_nx(G2)
net2.save_graph("networkx-pyvis-MCF2.html")
```

```{r print-HTML-2}
htmltools::renderDocument(htmltools::htmlTemplate("networkx-pyvis-MCF2.html"))
```
